---
title: "Correlation and Basic Regression"
subtitle: "SPS 502"
format: 
  revealjs:
    theme: dark
    slide-number: true
    preview-links: auto
---

## Correlation

- **Purpose:** Determine how closely two variables are related and the direction of the relationship
- Possible range from -1 to 1
  - -1 = perfect negative correlation
  - 1 = perfect positive correlation
  - 0 = perfectly uncorrelated

---

## Human Height and Weight

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
#| message: false
library(tidyverse)
library(patchwork)
set.seed(123)
n <- 1000

height_weight_data <- tibble(
  gender = rep(c("Female", "Male"), each = n),
  weight = c(rnorm(n, 170, 30), rnorm(n, 200, 25))
) %>%
  mutate(
    height = case_when(
      gender == "Female" ~ 55 + 0.08 * weight + rnorm(n(), 0, 2),
      gender == "Male" ~ 60 + 0.07 * weight + rnorm(n(), 0, 2)
    )
  )

correlations <- height_weight_data %>%
  group_by(gender) %>%
  summarize(correlation = cor(weight, height))

cor_female <- correlations %>% filter(gender == "Female") %>% pull(correlation)
cor_male <- correlations %>% filter(gender == "Male") %>% pull(correlation)
```

::: {.r-stack}
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6

#axis limit to make graphs same size
x_limits <- range(height_weight_data$weight)
y_limits <- range(height_weight_data$height)

x_label_female <- x_limits[1] + 0.7 * diff(x_limits)
y_label_female <- y_limits[1] + 0.1 * diff(y_limits)
x_label_male <- x_limits[1] + 0.7 * diff(x_limits)
y_label_male <- y_limits[1] + 0.1 * diff(y_limits)

# Step 1: Empty plots with axes only
p1_empty <- height_weight_data %>%
  filter(gender == "Female") %>%
  ggplot(aes(x = weight, y = height)) +
  labs(x = "Weight (pounds)", y = "Height (inches)") +
  facet_wrap(~gender) +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_dark(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p2_empty <- height_weight_data %>%
  filter(gender == "Male") %>%
  ggplot(aes(x = weight, y = height)) +
  labs(x = "Weight (pounds)", y = "Height (inches)") +
  facet_wrap(~gender) +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p1_empty + p2_empty
```

::: {.fragment}
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
# Step 2: Plots with scatterplots
p1_scatter <- height_weight_data %>%
  filter(gender == "Female") %>%
  ggplot(aes(x = weight, y = height)) +
  geom_point(color = "#FF9999", alpha = 0.5, size = 1) +
  labs(x = "Weight (pounds)", y = "Height (inches)") +
  facet_wrap(~gender) +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p2_scatter <- height_weight_data %>%
  filter(gender == "Male") %>%
  ggplot(aes(x = weight, y = height)) +
  geom_point(color = "#66CCCC", alpha = 0.5, size = 1) +
  labs(x = "Weight (pounds)", y = "Height (inches)") +
  facet_wrap(~gender) +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p1_scatter + p2_scatter
```
:::

::: {.fragment}
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
# Step 3: Plots with scatterplots and correlations
p1_final <- height_weight_data %>%
  filter(gender == "Female") %>%
  ggplot(aes(x = weight, y = height)) +
  geom_point(color = "#FF9999", alpha = 0.5, size = 1) +
  labs(x = "Weight (pounds)", y = "Height (inches)") +
  facet_wrap(~gender) +
  annotate("label", x = x_label_female, y = y_label_female, 
           label = paste0("r = ", round(cor_female, 3)),
           size = 8, fill = "#3399FF", color = "white") +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p2_final <- height_weight_data %>%
  filter(gender == "Male") %>%
  ggplot(aes(x = weight, y = height)) +
  geom_point(color = "#66CCCC", alpha = 0.5, size = 1) +
  labs(x = "Weight (pounds)", y = "Height (inches)") +
  facet_wrap(~gender) +
  annotate("label", x = x_label_male, y = y_label_male, 
           label = paste0("r = ", round(cor_male, 3)),
           size = 8, fill = "#3399FF", color = "white") +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p1_final + p2_final
```
:::
:::

::: {.fragment}
Strong _positive_ correlation
:::
---

## Car weight and fuel efficiency

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false
library(tidyverse)
set.seed(456)
n <- 40
car_data <- tibble(
  weight = runif(n, 1.5, 5.5),
  mpg = 35 - 4 * weight + rnorm(n, 0, 2)
)
cor_cars <- car_data %>%
  summarize(cor = cor(weight, mpg)) %>%
  pull(cor)

# Calculate axis limits once
x_limits <- range(car_data$weight)
y_limits <- range(car_data$mpg)

# Calculate label position within the plot area
x_label <- x_limits[1] + 0.7 * diff(x_limits)
y_label <- y_limits[1] + 0.1 * diff(y_limits)
```

::: {.r-stack}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
# Step 1: Empty plot with axes only
car_data %>%
  ggplot(aes(x = weight, y = mpg)) +
  labs(x = "Car weight (tons)", 
       y = "Miles per gallon") +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 16)
```

::: {.fragment}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
# Step 2: Plot with scatterplot
car_data %>%
  ggplot(aes(x = weight, y = mpg)) +
  geom_point(size = 3) +
  labs(x = "Car weight (tons)", 
       y = "Miles per gallon") +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 16)
```
:::

::: {.fragment}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
# Step 3: Plot with scatterplot AND correlation
car_data %>%
  ggplot(aes(x = weight, y = mpg)) +
  geom_point(size = 3) +
  labs(x = "Car weight (tons)", 
       y = "Miles per gallon") +
  annotate("label", x = x_label, y = y_label, 
           label = paste0("r = ", round(cor_cars, 3)),
           size = 8, fill = "#3399FF", color = "white") +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 16)
```
:::
:::

::: {.fragment}
Strong _negative_ correlation
:::

---

## IQ and Weight

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
#| message: false
library(tidyverse)
library(patchwork)
set.seed(789)
n <- 1500

# Create combined dataset
iq_data <- tibble(
  gender = rep(c("Female", "Male"), each = n),
  weight = c(rnorm(n, 170, 30), rnorm(n, 200, 30)),
  iq = rnorm(n * 2, 110, 15)
)

# Calculate correlations
correlations_iq <- iq_data %>%
  group_by(gender) %>%
  summarize(correlation = cor(weight, iq))

cor_female_iq <- correlations_iq %>% 
  filter(gender == "Female") %>% 
  pull(correlation)

cor_male_iq <- correlations_iq %>% 
  filter(gender == "Male") %>% 
  pull(correlation)

# Calculate axis limits once
x_limits <- range(iq_data$weight)
y_limits <- range(iq_data$iq)

# Calculate label positions within the plot area
x_label_female <- x_limits[1] + 0.7 * diff(x_limits)
y_label_female <- y_limits[1] + 0.1 * diff(y_limits)
x_label_male <- x_limits[1] + 0.7 * diff(x_limits)
y_label_male <- y_limits[1] + 0.1 * diff(y_limits)
```

::: {.r-stack}
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
# Step 1: Empty plots with axes only
p1_empty <- iq_data %>%
  filter(gender == "Female") %>%
  ggplot(aes(x = weight, y = iq)) +
  labs(x = "Weight (pounds)", y = "IQ") +
  facet_wrap(~gender) +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p2_empty <- iq_data %>%
  filter(gender == "Male") %>%
  ggplot(aes(x = weight, y = iq)) +
  labs(x = "Weight (pounds)", y = "IQ") +
  facet_wrap(~gender) +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p1_empty + p2_empty
```

::: {.fragment}
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
# Step 2: Plots with scatterplots
p1_scatter <- iq_data %>%
  filter(gender == "Female") %>%
  ggplot(aes(x = weight, y = iq)) +
  geom_point(color = "#FF9999", alpha = 0.4, size = 1) +
  labs(x = "Weight (pounds)", y = "IQ") +
  facet_wrap(~gender) +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p2_scatter <- iq_data %>%
  filter(gender == "Male") %>%
  ggplot(aes(x = weight, y = iq)) +
  geom_point(color = "#66CCCC", alpha = 0.4, size = 1) +
  labs(x = "Weight (pounds)", y = "IQ") +
  facet_wrap(~gender) +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p1_scatter + p2_scatter
```
:::

::: {.fragment}
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
# Step 3: Plots with scatterplots AND correlations
p1_final <- iq_data %>%
  filter(gender == "Female") %>%
  ggplot(aes(x = weight, y = iq)) +
  geom_point(color = "#FF9999", alpha = 0.4, size = 1) +
  labs(x = "Weight (pounds)", y = "IQ") +
  facet_wrap(~gender) +
  annotate("label", x = x_label_female, y = y_label_female, 
           label = paste0("r = ", round(cor_female_iq, 3)),
           size = 8, fill = "#3399FF", color = "white") +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p2_final <- iq_data %>%
  filter(gender == "Male") %>%
  ggplot(aes(x = weight, y = iq)) +
  geom_point(color = "#66CCCC", alpha = 0.4, size = 1) +
  labs(x = "Weight (pounds)", y = "IQ") +
  facet_wrap(~gender) +
  annotate("label", x = x_label_male, y = y_label_male, 
           label = paste0("r = ", round(cor_male_iq, 3)),
           size = 8, fill = "#3399FF", color = "white") +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(size = 16))

p1_final + p2_final
```
:::
:::

::: {.fragment}
**NO** correlation
:::

---

## General Guidelines

| Range | Interpretation | 
|-------|----------------|
| 0 | No relationship |
| 0.01–0.19 | Little to no relationship |
| 0.20–0.29 | Weak relationship |
| 0.30–0.39 | Moderate relationship |
| 0.40–0.69 | Strong relationship |
| 0.70–0.99 | Very strong relationship |
| 1 | Perfect relationship |

::: {.callout-note}
Can be positive or negative
:::

---

## Interpreting Correlation Coefficients

> As the value of X goes up, Y tends to go up (or down) a lot/a little/not at all

---


## Guess the Correlation

::: columns
::: {.column width="50%"}
**X** 

::: {.fragment fragment-index=2}
Vehicle velocity
:::

::: {.fragment fragment-index=4}
Time in School
:::

::: {.fragment fragment-index=6}
Alcohol consumed
:::

::: {.fragment fragment-index=8}
Income
:::

::: {.fragment fragment-index=10}
Age
:::

::: {.fragment fragment-index=12}
Hair length
:::

::: {.fragment fragment-index=14}
Yard Size
:::

:::

::: {.column width="50%"}

**Y**

::: {.fragment fragment-index=1}
Travel time
:::

::: {.fragment fragment-index=3}
Debt
:::

::: {.fragment fragment-index=5}
Judgment
:::

::: {.fragment fragment-index=7}
Happiness
:::

::: {.fragment fragment-index=9}
Health
:::

::: {.fragment fragment-index=11}
Shampoo use
:::

::: {.fragment fragment-index=13}
Water bill
:::

:::
:::

## Basic Regression

- Correlation only tells part of the story (direction and magnitude of relationship)
- Regression allows us to **predict** events and/or **explain** why they occur


## Essential Parts

::: {.columns}
::: {.column width="50%"}
### Y

- Outcome variable
- Response variable
- Dependent variable
- Thing you want to explain or predict
:::

::: {.column width="50%"}
### ~X (or lots of Xs)

- Explanatory variable
- Predictor variable
- Independent variable
- Thing you use to explain changes in Y
:::
:::


## Identify the Y and X Variables

::: {.fragment .fade-in}
::: {.fragment .semi-fade-out}
A study examines the effect of smoking on lung cancer
:::
:::

::: {.fragment .fade-in}
::: {.fragment .semi-fade-out}
Researchers predict genocides by looking at negative media coverage, revolutions in neighboring countries, and economic growth
:::
:::

::: {.fragment .fade-in}
::: {.fragment .semi-fade-out}
You want to see if students taking more AP classes in high school improves their college grades
:::
:::

::: {.fragment .fade-in}
::: {.fragment .semi-fade-out}
Netflix uses your past viewing history, the day of the week, and the time of the day to guess which show you want to watch next
:::
:::


## Two Purposes of Regression

::: {.columns}
::: {.column width="50%"}
::: {.callout-note icon="false"}
## 📊 Prediction

::: {.incremental}
- **Forecast the future**
- **Focus is on Y**
:::

::: {.fragment}
##### Examples:
:::

::: {.fragment}
::: {.callout-tip appearance="minimal"}
Netflix trying to guess your next show
:::
:::

::: {.fragment}
::: {.callout-tip appearance="minimal"}
Predicting who will escape poverty
:::
:::
:::
:::

::: {.column width="50%"}
::: {.callout-note icon="false"}
## 🔍 Explanation

::: {.incremental}
- **Explain effect of X on Y**
- **Focus is on X**
:::

::: {.fragment}
##### Examples:
:::

::: {.fragment}
::: {.callout-tip appearance="minimal"}
Netflix estimating the effect of time of day on show selection
:::
:::

::: {.fragment}
::: {.callout-tip appearance="minimal"}
Estimating the effect of food stamps on poverty reduction
:::
:::
:::
:::
:::

---

## Regression Process

::: {.callout-important icon="false"}
## How to Perform Regression

:::: {.incremental}
1. **Plot X and Y**  
   *Visualize the relationship between variables*

2. **Draw a line that approximates the relationship**  
   *Find the best fit through the data points*

3. **Find the "mathy" parts of the line**  
   *Calculate slope and intercept*

4. **Interpret the math**  
   *Translate coefficients into meaningful insights*
::::
:::

## Cookies and Happiness Data

```{r}
#| echo: true

library(tidyverse)

cookies_data <- tibble(
  happiness = c(0.5, 2.0, 1.0, 2.5, 3.0, 1.5, 2.0, 2.5, 2.0, 3.0),
  cookies = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
)

cookies_data
```

---

## Scatterplot: Cookies and Happiness

::: {.r-stack}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false
library(tidyverse)

# Base plot without points
cookies_data %>%
  ggplot(aes(x = cookies, y = happiness)) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Relationship between cookies and happiness") +
  theme_minimal(base_size = 16) +
  scale_y_continuous(limits = c(0, 3))
```

::: {.fragment}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false

# Add points
cookies_data %>%
  ggplot(aes(x = cookies, y = happiness)) +
  geom_point(size = 4) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Relationship between cookies and happiness") +
  theme_minimal(base_size = 16) +
  scale_y_continuous(limits = c(0, 3))
```
:::


::: {.fragment}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false

# Add correlation coefficient
cor_cookies <- cookies_data %>%
  summarize(cor = cor(cookies, happiness)) %>%
  pull(cor)

cookies_data %>%
  ggplot(aes(x = cookies, y = happiness)) +
  geom_point(size = 4) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Relationship between cookies and happiness") +
  annotate("label", x = 8.5, y = 0.7, 
           label = paste0("r = ", round(cor_cookies, 3)),
           size = 8, fill = "#3399FF", color = "white") +
  theme_minimal(base_size = 16) +
  scale_y_continuous(limits = c(0, 3))
```
:::
:::

## Finding the Best Line

::: {.r-stack}
::: {.fragment}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false


cookies_data %>%
  ggplot(aes(x = cookies, y = happiness)) +
  geom_line(color = "#3399FF", se = FALSE, linewidth = 1.5) +
  geom_point(size = 4) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Relationship between cookies and happiness") +
  theme_minimal(base_size = 16) +
  scale_y_continuous(limits = c(0, 3))
```
:::


::: {.fragment}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false


cookies_data %>%
  ggplot(aes(x = cookies, y = happiness)) +
  geom_smooth(color = "#3399FF", se = FALSE, linewidth = 1.5) +
  geom_point(size = 4) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Relationship between cookies and happiness") +
  theme_minimal(base_size = 16) +
  scale_y_continuous(limits = c(0, 3))
```
:::

::: {.fragment}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false

library(tidyverse)

cookies_data %>%
  ggplot(aes(x = cookies, y = happiness)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, color = "#3399FF", linewidth = 1.5) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Relationship between cookies and happiness") +
  theme_minimal(base_size = 16) +
  scale_y_continuous(limits = c(0, 3))
```
:::
:::

---

## Residual Errors

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false

library(tidyverse)
library(broom)

model <- lm(happiness ~ cookies, data = cookies_data)

cookies_augmented <- cookies_data %>%
  mutate(
    predicted = predict(model),
    residuals = residuals(model)
  )

cookies_augmented %>%
  ggplot(aes(x = cookies, y = happiness)) +
  geom_smooth(method = "lm", se = FALSE, color = "#3399FF", linewidth = 1.5) +
  geom_segment(aes(xend = cookies, yend = predicted), 
               color = "#FF9900", linewidth = 1) +
  geom_point(size = 4) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Relationship between cookies and happiness") +
  theme_minimal(base_size = 16) +
  scale_y_continuous(limits = c(0, 3))
```

---

## Residual Errors (Distance from Line)

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false

library(tidyverse)

cookies_augmented %>%
  ggplot(aes(x = cookies, y = residuals)) +
  geom_hline(yintercept = 0, color = "purple", linewidth = 1.5) +
  geom_segment(aes(xend = cookies, yend = 0), 
               color = "#FF9900", linewidth = 1) +
  geom_point(size = 4, color = "#FF9900") +
  labs(x = "Cookies eaten", 
       y = "Distance from line",
       title = "Residual errors (distance from line)") +
  theme_minimal(base_size = 16)
```

---

## 

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6
#| message: false

library(tidyverse)
library(patchwork)

p1 <- cookies_augmented %>%
  ggplot(aes(x = cookies, y = happiness)) +
  geom_smooth(method = "lm", se = FALSE, color = "#3399FF", linewidth = 1.5) +
  geom_segment(aes(xend = cookies, yend = predicted), 
               color = "#FF9900", linewidth = 1) +
  geom_point(size = 3) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Cookies and happiness") +
  theme_minimal(base_size = 12) +
  scale_y_continuous(limits = c(0, 3))

p2 <- cookies_augmented %>%
  ggplot(aes(x = cookies, y = residuals)) +
  geom_hline(yintercept = 0, color = "purple", linewidth = 1.5) +
  geom_segment(aes(xend = cookies, yend = 0), 
               color = "#FF9900", linewidth = 1) +
  geom_point(size = 3, color = "#FF9900") +
  labs(x = "Cookies eaten", 
       y = "Distance from line",
       title = "Residual errors") +
  theme_minimal(base_size = 12)

p1 + p2
```

---



## Ordinary Least Squares (OLS) Regression

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false

library(tidyverse)

cookies_data %>%
  ggplot(aes(x = cookies, y = happiness)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, color = "#3399FF", linewidth = 1.5) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Relationship between cookies and happiness") +
  annotate("label", x = 7.5, y = 1, 
           label = "Ordinary least\nsquares (OLS)\nregression",
           size = 6, fill = "#66CCFF", color = "white") +
  theme_minimal(base_size = 16) +
  scale_y_continuous(limits = c(0, 3))
```

---

## The Math Part

$$y = mx + b$$

::: {.incremental}
- **y**: A number
- **x**: A number
- **m**: Slope
- **b**: y intercept
:::

---

## Slopes and Intercepts

::: {.columns}
::: {.column width="50%"}
### y = 2x - 1

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 5
#| message: false

library(tidyverse)

tibble(x = 0:5) %>%
  mutate(y = 2*x - 1) %>%
  ggplot(aes(x, y)) +
  geom_line(color = "darkblue", linewidth = 2) +
  theme_minimal(base_size = 14) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(breaks = seq(0, 9, 1))
```
:::

::: {.column width="50%"}
### y = -0.5x + 6

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 5
#| message: false

library(tidyverse)

tibble(x = 0:14) %>%
  mutate(y = -0.5*x + 6) %>%
  ggplot(aes(x, y)) +
  geom_line(color = "darkblue", linewidth = 2) +
  theme_minimal(base_size = 14) +
  scale_x_continuous(breaks = seq(0, 14, 1))
```

:::
:::

---

## Drawing Lines with Stats

$$\widehat{y} = \beta_0 + \beta_1 x + \varepsilon$$

::: {.incremental}
- $\widehat{y}$: Outcome variable
- $x$: Explanatory variable  
- $\beta_1$: Slope
- $\beta_0$: y intercept
- $\varepsilon$: Error (residuals)
:::

---

## Modeling Cookies and Happiness

$\widehat{y} = \beta_0 + \beta_1 x_1 + \varepsilon$

$\widehat{happiness} = \beta_0 + \beta_1 \times cookies + \varepsilon$

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| message: false

library(tidyverse)

cookies_data %>%
  ggplot(aes(x = cookies, y = happiness)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, color = "#3399FF", linewidth = 1.5) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Relationship between cookies and happiness") +
  theme_minimal(base_size = 14) +
  scale_y_continuous(limits = c(0, 3))
```

---

## Modeling Cookies & Happiness in R

```{r}
#| echo: true
#| eval: false

cookies_model <- lm(happiness ~ cookies, 
                    data = cookies_data)

cookies_model %>%
  get_regression_table()
```

```{r}
#| echo: false

library(moderndive)

cookies_model <- lm(happiness ~ cookies, data = cookies_data)
get_regression_table(cookies_model)
```

---

## Modeling Cookies & Happiness in R

$$\widehat{happiness} = 1.1 + 0.164 \times cookies + \varepsilon$$

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5

ggplot(cookies_data, aes(x = cookies, y = happiness)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, color = "#3399FF", linewidth = 1.5) +
  labs(x = "Cookies eaten", 
       y = "Level of happiness",
       title = "Relationship between cookies and happiness") +
  theme_minimal(base_size = 14) +
  scale_y_continuous(limits = c(0, 3))
```

| term | estimate | std_error | statistic | p_value | lower_ci | upper_ci |
|------|----------|-----------|-----------|---------|----------|----------|
| intercept | 1.1 | 0.47 | 2.339 | 0.047 | 0.016 | 2.184 |
| cookies | 0.164 | 0.076 | 2.159 | 0.063 | -0.011 | 0.338 |

---

## Interpreting Regression Coefficients
::: {.fragment}
> A one unit increase in X is associated with a $\beta_1$ increase (or decrease) in Y, on average

::: 

::: {.fragment}
$$\widehat{happiness} = 1.1 + 0.164 \times cookies + \varepsilon$$
:::

::: {.fragment}
$$\widehat{y} = \beta_0 + \beta_1 x_1 + \varepsilon$$
:::

::: {.fragment}
**Interpretation:** A one-cookie increase is associated with a 0.164 point increase in happiness, on average.
::: 