---
title: "Multiple Regression and Diagnostics"
subtitle: "SPS 502"
format:
  revealjs:
    slide-number: true
    code-overflow: wrap
    self-contained: true
    toc: false
execute:
  echo: true
  warning: false
  message: false
---

> Slides mirror the provided deck except Homework 2 and Exam 1. Data: `/data/world_happiness.csv`.

## Setup

```{r}
#| label: setup
#| include: false
packages <- c("tidyverse", "broom", "forcats", "gt")
need <- setdiff(packages, rownames(installed.packages()))
if (length(need)) install.packages(need, quiet = TRUE)
invisible(lapply(packages, library, character.only = TRUE))

# Load and clean data
wh <- read_csv("data/world_happiness.csv", show_col_types = FALSE)

# Clean column names using base R instead of janitor
names(wh) <- tolower(gsub("[^a-zA-Z0-9]", "_", names(wh)))
names(wh) <- gsub("_+", "_", names(wh))  # Remove multiple underscores
names(wh) <- gsub("^_|_$", "", names(wh))  # Remove leading/trailing underscores

# Check data structure and available variables
has_school <- "school_enrollment" %in% names(wh)
has_region <- "region" %in% names(wh)

# Clean and factor region variable if it exists
if (has_region) {
  wh <- wh %>% 
    filter(!is.na(region)) %>%
    mutate(region = str_trim(region),
           region = forcats::fct_relevel(region, "East Asia & Pacific"))
}

# Remove rows with missing happiness scores
wh <- wh %>% filter(!is.na(happiness_score))
```

## Goal of Regression

Explain or predict variation in an outcome using one or more predictors.

## Algebra → Stats

$$\widehat{y} = \beta_0 + \beta_1 x + \varepsilon$$

::: {.incremental}
- $\widehat{y}$: Outcome variable
- $x$: Explanatory variable  
- $\beta_1$: Slope
- $\beta_0$: y intercept
- $\varepsilon$: Error (residuals)
:::


## Example: Cookies and Happiness

```{r}
#| echo: false
# toy data to illustrate simple regression
set.seed(123)
cookies_df <- tibble(
  cookies = 1:10,
  happiness = 1.1 + 0.164 * cookies + rnorm(10, sd = 0.35)
)

mod_cookies <- lm(happiness ~ cookies, data = cookies_df)
broom::tidy(mod_cookies, conf.int = TRUE) %>%
  gt() %>%
  fmt_number(columns = c(estimate, std.error, statistic, p.value, conf.low, conf.high), decimals = 3)
```

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
ggplot(cookies_df, aes(cookies, happiness)) +
  geom_point(size = 3, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Relationship between cookies and happiness",
       x = "Cookies eaten", y = "Level of happiness") +
  theme_minimal()
```

**Interpretation of slope.** A 1-unit increase in `cookies` is associated with a $\beta_1$ change in `happiness`, on average.

## Variable Types

- Numeric (continuous)  
- Categorical (factors)

## Life Expectancy and Happiness

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 6
if (has_region) {
  ggplot(wh, aes(life_expectancy, happiness_score, color = region)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(x = "Life expectancy", y = "Happiness score",
         title = "Life Expectancy vs Happiness by Region") +
    theme_minimal() +
    theme(legend.position = "bottom")
} else {
  ggplot(wh, aes(life_expectancy, happiness_score)) +
    geom_point(alpha = 0.7, size = 2, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(x = "Life expectancy", y = "Happiness score",
         title = "Life Expectancy vs Happiness") +
    theme_minimal()
}
```

## Regional Differences

```{r}
#| echo: false
if (has_region) {
  region_avg <- wh %>%
    group_by(region) %>%
    summarise(
      avg_happiness = mean(happiness_score, na.rm = TRUE),
      n_countries = n(),
      .groups = "drop"
    ) %>%
    arrange(desc(avg_happiness))
  
  region_avg %>% 
    gt() %>%
    cols_label(
      region = "Region",
      avg_happiness = "Average Happiness",
      n_countries = "Number of Countries"
    ) %>%
    fmt_number(columns = avg_happiness, decimals = 2)
} else {
  tibble(note = "Region data not available in this dataset") %>% gt()
}
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
if (has_region && exists("region_avg")) {
  ggplot(region_avg, aes(reorder(region, avg_happiness), avg_happiness)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    coord_flip() +
    labs(x = NULL, y = "Average happiness",
         title = "Average Happiness by Region") +
    theme_minimal()
}
```

## Model: Happiness ~ Region

```{r}
#| echo: false
if (has_region) {
  mod_region <- lm(happiness_score ~ region, data = wh)
  broom::tidy(mod_region, conf.int = TRUE) %>%
    mutate(term = str_replace(term, "^region", "region: ")) %>%
    gt() %>%
    cols_label(
      term = "Term",
      estimate = "Estimate",
      std.error = "Std. Error",
      statistic = "t-statistic",
      p.value = "p-value",
      conf.low = "95% CI Lower",
      conf.high = "95% CI Upper"
    ) %>%
    fmt_number(columns = c(estimate, std.error, statistic, conf.low, conf.high), decimals = 3) %>%
    fmt_scientific(columns = p.value, decimals = 3)
} else {
  tibble(note = "Cannot run region model - region data not available") %>% gt()
}
```

## Predicted Means by Region (baseline = East Asia & Pacific)

```{r}
#| echo: false
if (has_region && exists("mod_region")) {
  co <- coef(mod_region)
  baseline <- unname(co[["(Intercept)"]])
  
  # Get all region coefficients
  region_coefs <- co[grepl("^region", names(co))]
  
  # Create predicted means table
  pred_means <- tibble(
    region = c("East Asia & Pacific", str_replace(names(region_coefs), "^region", "")),
    predicted_mean = c(baseline, baseline + unname(region_coefs))
  ) %>%
    arrange(desc(predicted_mean))
  
  pred_means %>%
    gt() %>%
    cols_label(
      region = "Region",
      predicted_mean = "Predicted Mean Happiness"
    ) %>%
    fmt_number(columns = predicted_mean, decimals = 3)
}
```

- Compared to East Asia & Pacific, the coefficient on each region is the mean difference in happiness.

## Multiple Regression

```{r}
#| echo: false
if (!has_school) {
  cat("Note: 'school_enrollment' not found in dataset. Using available predictors instead.\n")
  # Show available numeric variables
  numeric_vars <- wh %>% select(where(is.numeric)) %>% names()
  cat("Available numeric variables:", paste(numeric_vars, collapse = ", "), "\n")
}
```

```{r}
#| echo: false
if (has_school) {
  mod_life_school <- lm(happiness_score ~ life_expectancy + school_enrollment, data = wh)
  broom::tidy(mod_life_school, conf.int = TRUE) %>% 
    gt() %>%
    cols_label(
      term = "Term",
      estimate = "Estimate",
      std.error = "Std. Error",
      statistic = "t-statistic",
      p.value = "p-value",
      conf.low = "95% CI Lower",
      conf.high = "95% CI Upper"
    ) %>%
    fmt_number(columns = c(estimate, std.error, statistic, conf.low, conf.high), decimals = 3) %>%
    fmt_scientific(columns = p.value, decimals = 3)
} else {
  # Use an alternative predictor if school enrollment isn't available
  alt_vars <- setdiff(names(wh %>% select(where(is.numeric))), c("happiness_score", "year"))
  if (length(alt_vars) >= 1) {
    alt_var <- alt_vars[1]  # Use first available numeric variable
    mod_alt <- lm(as.formula(paste("happiness_score ~ life_expectancy +", alt_var)), data = wh)
    broom::tidy(mod_alt, conf.int = TRUE) %>% 
      gt() %>%
      cols_label(
        term = "Term",
        estimate = "Estimate",
        std.error = "Std. Error",
        statistic = "t-statistic",
        p.value = "p-value",
        conf.low = "95% CI Lower",
        conf.high = "95% CI Upper"
      ) %>%
      fmt_number(columns = c(estimate, std.error, statistic, conf.low, conf.high), decimals = 3) %>%
      fmt_scientific(columns = p.value, decimals = 3)
  }
}
```

**Interpretation.** Holding other variables in the model constant, a 1-unit increase in a predictor is associated with a \(\beta\) change in the outcome, on average.

## R² and Adjusted R²

How much variation in \(y\) is explained by \(x\). Scale 0–1. Higher indicates better fit.

Adjusted \(R^2\) penalizes adding predictors relative to sample size:  
\(
R^2_{\text{adj}} = 1 - (1 - R^2)\frac{n - 1}{n - p - 1}
\)

## Finding R² in R

```{r}
#| echo: false
model1 <- lm(happiness_score ~ life_expectancy, data = wh)
broom::glance(model1) %>% 
  select(r.squared, adj.r.squared) %>% 
  gt() %>%
  cols_label(
    r.squared = "R²",
    adj.r.squared = "Adjusted R²"
  ) %>%
  fmt_number(columns = c(r.squared, adj.r.squared), decimals = 3)
```

```{r}
#| echo: false
if (has_school && exists("mod_life_school")) {
  broom::glance(mod_life_school) %>% 
    select(r.squared, adj.r.squared) %>% 
    gt() %>%
    cols_label(
      r.squared = "R²",
      adj.r.squared = "Adjusted R²"
    ) %>%
    fmt_number(columns = c(r.squared, adj.r.squared), decimals = 3)
} else if (exists("mod_alt")) {
  broom::glance(mod_alt) %>% 
    select(r.squared, adj.r.squared) %>% 
    gt() %>%
    cols_label(
      r.squared = "R²",
      adj.r.squared = "Adjusted R²"
    ) %>%
    fmt_number(columns = c(r.squared, adj.r.squared), decimals = 3)
}
```

## Interpreting R²

- “This model explains X% of the variation in the outcome.”  
- Use \(R^2\) for one predictor. Use adjusted \(R^2\) for multiple predictors.  
- Higher is better. No universal threshold.
